---
title: "Kidz Bop Censored Lyrics"
description: |
  Analysis of censored lyrics in Kidz Bop songs
author: "Guy Fuchs"
date: today
format: html
execute:
  warning: false
  message: false
---

> Dataset page: [The Pudding – Kidz Bop Data](https://github.com/the-pudding/data/tree/master/kidz-bop)  
> Original source: [Kidz Bop Censored Lyrics Dataset (Hehmeyer et al., The Pudding)](https://raw.githubusercontent.com/the-pudding/data/master/kidz-bop/KB_censored-lyrics.csv)

```{r}
#| label: setup
#| code-fold: true
library(tidyverse)
library(scales)
```

```{r}
kb <- read_csv("https://raw.githubusercontent.com/the-pudding/data/master/kidz-bop/KB_censored-lyrics.csv")

category_counts <- kb |>
  count(category) |>
  head(10)

top8 <- head(category_counts, 8) 

ggplot(data = top8, aes(x = reorder(category, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "What Kidz Bop censors most",
    subtitle = "Top categories in the censored-lyrics dataset",
    x = "Category",
    y = "Censored instances"
  ) +
  theme_minimal()
```

This plot helps to show which types of content are most frequently censored in Kidz Bop songs. From the counts, we can see that profanity and sexual references make up the largest portion of censored lyrics, followed closely by alcohol and drug references. Categories like violence, identity, and other themes appear much less often.

These patterns reveal that Kidz Bop's censorship choices focus mainly on removing language that could be considered inappropriate or too mature for a younger audience, while other categories are less frequently flagged. This highlights how producers are likely prioritizing a "family-friendly" tone in their songs.

```{r}
kb|>
  mutate(my_word=str_extract_all(ogLyric, "(?<=\\bmy\\s)\\w+")) |>
  select(my_word) |>
  unnest(my_word) |>
  count(my_word, sort = TRUE) |>
  head(10)
```

This table shows which words most often follow the word "my" in the original lyrics of the Kidz Bop songs. We can see that the top three words that follow "my" are nights, woman, and God, highlighting the personal and relational themes in censored lyrics.

```{r}
kb |>
  mutate(first_three_words = str_extract(ogLyric, "^(?:\\S+\\s+){0,2}\\S+")) |>  
  count(first_three_words, sort = TRUE) |>
  head(10)
```

This table shows the most common three-word openings found in the original lyrics of Kidz Bop songs. We can see that many of the top phrases like "Blow a kiss", "Be my woman", and "Kissed her on" start with actions and expressions of relationships. These repeated openings show that censored lyrics often begin with personal and romantic themes that would likely need to be toned down for the younger audience of Kidz Bop. 


```{r}

kb_lower <- kb |>
  filter(!is.na(year), !is.na(ogLyric))|>
  mutate(og_lower = str_to_lower(ogLyric))

reg_expression <- "\\b(love|lover|kiss|kissing|baby|girlfriend|boyfriend|girl|boy|wife|husband|ex|partner)\\b"

by_year <- kb_lower |>
  mutate(has_relationship = str_detect(og_lower, reg_expression)) |>
  group_by(year) |>
  summarise(total = n(), relationship = sum(has_relationship, na.rm = TRUE), pct = relationship / total) |>
  filter(total >= 20) |> # helps to stabilize the percentages
  ungroup()

ggplot(by_year, aes(x = year, y = pct)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Lyrics mentioning relationships over time",
    subtitle = "Share of original lyrics with partner or family terms (years with 20+ rows)",
    x = "Year",
    y = "Percent of lyrics mentioning relationships"
  ) +
  theme_minimal()
```

This plots shows how often relationship and love-related words such as love, kiss, baby, girlfriend, boyfriend, etc. appear in the original Kidz Bop lyrics over time. Although the percentages remain small overall, there are clear ups and downs across the years. At peaks around 2014-2015, the data suggests that more popular songs from those years contained romantic and affectionate language that Kidz Bop chose to censor and modify. On the other hand, lower points indicate years when songs with relationship themes were a little bit less common. Overall, we can see how the romantic tone of pop music can change over time and how Kidz Bop's censorship accommodates those changing tones.

